---
title: "smFISH PDF-mask summary"
subtitle: "Neurite vs soma spot counts"
author: "Bader project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    theme: cosmo
    df-print: paged
params:
  summary_csv: "pdf_mask_summary.csv"
  use_dummy_data: true
  # 6-digit leading dates are ambiguous -- "dmy" reads 130323 as 13 Mar 2023
  # (the convention in the existing lab filenames), "ymd" as 2013-03-23.
  date_format: "dmy"
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
```

## What this does

Reads the summary CSV written by `count_spots_in_pdf_mask.py`, which has just two
columns:

| column | meaning |
|--------|---------|
| `csv_filename` | name of the bigfish spot CSV, e.g. `260804_Camk2a_neurite_1_ch3_spots.csv` |
| `nb_spots_within_mask` | spots falling inside the PDF mask |

Everything else — date, gene, compartment — is encoded in the filename, so this
document parses it back out, then pivots soma and neurite onto one row per
sample to compute `neurite_soma_ratio`.

::: callout-warning
## Dummy filename convention — edit before real use

This is a **scaffold**. The parsing below assumes:

```
{date}_{gene}_{compartment}_{replicate}_ch{N}_spots.csv
040826_Camk2a_neurite_1_ch3_spots.csv
```

Your real filenames almost certainly differ, and the current pipeline does not
encode compartment at all — `count_spots_in_pdf_mask.py` writes whatever the
microscope file was called, plus `_ch{N}_spots.csv`. The three `case_when()`
blocks in [Parse the filename] are the only place to change; each has an
explicit `.default` marking unmatched files `NA` rather than guessing, and
[Parsing QC] stops the render if anything comes back `NA`.

Two decisions I could not make for you:

-   **Date format.** A leading `130323` is ambiguous. `params$date_format`
    defaults to `"dmy"` (13 Mar 2023), matching the convention in your existing
    filenames like `130323_Dhc8-1-6-6_Gdi565_Rab11647_L2_1`. Set it to `"ymd"`
    if that reading is wrong.
-   **Gene list.** Hard-coded as `Camk2a`/`Bdnf`/`Actb` placeholders. Your real
    gene tokens (`Gdi565`, `Rab11647`, …) look like gene+fluorophore pairs and
    may need splitting into two columns.
:::

## Load

With `params$use_dummy_data: true` this fabricates a small example table so the
document always renders. Set it to `false` (or run
`quarto render ... -P use_dummy_data:false`) to read the real CSV.

```{r}
#| label: load

make_dummy_summary <- function() {
  tidyr::expand_grid(
    # DDMMYY, matching params$date_format: 040826 = 4 Aug 2026
    date       = c("040826", "050826"),
    gene       = c("Camk2a", "Bdnf", "Actb"),
    compartment = c("soma", "neurite"),
    replicate  = 1:3
  ) |>
    mutate(
      csv_filename = str_glue(
        "{date}_{gene}_{compartment}_{replicate}_ch3_spots.csv"
      ) |> as.character(),
      # soma-biased counts, with neurite deliberately lower
      nb_spots_within_mask = if_else(
        compartment == "soma",
        as.integer(round(runif(n(), 400, 900))),
        as.integer(round(runif(n(),  40, 300)))
      )
    ) |>
    select(csv_filename, nb_spots_within_mask)
}

raw <- if (isTRUE(params$use_dummy_data)) {
  set.seed(42)
  make_dummy_summary()
} else {
  readr::read_csv(params$summary_csv, show_col_types = FALSE)
}

stopifnot(all(c("csv_filename", "nb_spots_within_mask") %in% names(raw)))

glimpse(raw)
```

## Parse the filename

Three `case_when()` blocks, one per parameter. `str_detect()` with word
boundaries (`\\b`) rather than bare substrings, so a gene called `Actb` does not
also match a file containing `Actbl2`.

```{r}
#| label: parse

# Token delimiter: filename fields are separated by "_", and "_" is a word
# character, so \\b does NOT mark a boundary inside 260804_Camk2a_soma_1.
# Match on the delimiters explicitly instead.
# The (?:...) wrapper matters: without it, tok("soma|cell_body") would parse as
# "(^|_)soma" OR "cell_body(_|\\.|$)" and match in the wrong places.
tok <- function(x) regex(str_c("(^|_)(?:", x, ")(_|\\.|$)"), ignore_case = TRUE)

annotated <- raw |>
  mutate(
    # ---- date ------------------------------------------------------------
    # 6-digit dates are ambiguous; params$date_format decides. 8 digits is
    # unambiguous YYYYMMDD.
    date_raw = str_extract(csv_filename, "^\\d+"),
    date = case_when(
      str_length(date_raw) == 8                              ~ lubridate::ymd(date_raw, quiet = TRUE),
      str_length(date_raw) == 6 & params$date_format == "dmy" ~ lubridate::dmy(date_raw, quiet = TRUE),
      str_length(date_raw) == 6 & params$date_format == "ymd" ~ lubridate::ymd(date_raw, quiet = TRUE),
      .default = as.Date(NA)
    ),

    # ---- gene ------------------------------------------------------------
    gene = case_when(
      str_detect(csv_filename, tok("Camk2a")) ~ "Camk2a",
      str_detect(csv_filename, tok("Bdnf"))   ~ "Bdnf",
      str_detect(csv_filename, tok("Actb"))   ~ "Actb",
      .default = NA_character_
    ),

    # ---- compartment -----------------------------------------------------
    compartment = case_when(
      str_detect(csv_filename, tok("neurite|axon|dendrite")) ~ "neurite",
      str_detect(csv_filename, tok("soma|cell.?body"))       ~ "soma",
      .default = NA_character_
    ),

    # ---- replicate and channel ------------------------------------------
    replicate = str_extract(csv_filename, "_(\\d+)_ch\\d+_", group = 1),
    channel   = str_extract(csv_filename, "_ch(\\d+)_",      group = 1)
  ) |>
  select(-date_raw) |>
  relocate(date, gene, compartment, replicate, channel, .after = csv_filename)

annotated
```

### Parsing QC

This is a **hard stop, not a warning**. An `NA` gene does not merely drop a
sample — it makes every unparsed file share the same `id_cols` group, and
`values_fn = sum` in the pivot below then adds unrelated genes together into a
single row. The result looks entirely plausible and is wrong. Better to fail.

```{r}
#| label: parse-qc
#| error: true

unparsed <- annotated |>
  filter(if_any(c(date, gene, compartment), is.na))

if (nrow(unparsed) > 0) {
  print(unparsed |> select(csv_filename, date, gene, compartment), n = 20)
  stop(
    nrow(unparsed), " of ", nrow(annotated),
    " filename(s) did not parse. Fix the case_when() blocks above before ",
    "trusting anything below -- unparsed rows corrupt the pivot silently."
  )
}

cat("All", nrow(annotated), "filenames parsed cleanly.\n")
```

## Pivot to one row per sample

`pivot_wider()` needs to know what makes two rows *the same sample* — here
date + gene + replicate — so the soma and neurite rows collapse together.

`values_fn = sum` handles a sample that legitimately has more than one CSV per
compartment (e.g. two channels quantified separately). Without it
`pivot_wider()` returns list-columns and the ratio arithmetic breaks. But
summing is only correct when you *meant* to pool those files, so the function
reports every cell where it actually fired rather than pooling silently — if
you did not expect it, your `id_cols` are too coarse.

```{r}
#| label: widen

widen_by_compartment <- function(data,
                                 id_cols = c(date, gene, replicate),
                                 min_soma_count = 1) {

  # surface any (sample x compartment) cell built from more than one file
  pooled <- data |>
    filter(!is.na(compartment)) |>
    summarise(n_files = n(), .by = c({{ id_cols }}, compartment)) |>
    filter(n_files > 1)

  if (nrow(pooled) > 0) {
    warning(
      nrow(pooled), " cell(s) pooled from >1 file -- inspect with ",
      "attr(wide, 'pooled_cells')"
    )
  }

  pivoted <- data |>
    filter(!is.na(compartment)) |>
    pivot_wider(
      id_cols     = {{ id_cols }},
      names_from  = compartment,
      values_from = nb_spots_within_mask,
      names_glue  = "{compartment}_count",
      values_fn   = sum,
      values_fill = 0L
    )

  # tolerate a batch where one compartment is entirely absent
  for (col in c("soma_count", "neurite_count")) {
    if (!col %in% names(pivoted)) pivoted[[col]] <- NA_integer_
  }

  result <- pivoted |>
    mutate(
      total_count = soma_count + neurite_count,
      # guard division by zero: a soma with no spots gives no meaningful ratio
      neurite_soma_ratio = if_else(
        soma_count >= min_soma_count,
        neurite_count / soma_count,
        NA_real_
      ),
      neurite_fraction = if_else(
        total_count > 0,
        neurite_count / total_count,
        NA_real_
      )
    ) |>
    relocate(soma_count, neurite_count, total_count, .after = replicate) |>
    arrange(date, gene, replicate)

  attr(result, "pooled_cells") <- pooled
  result
}

wide <- widen_by_compartment(annotated)
wide
```

## Results

```{r}
#| label: table

if (requireNamespace("DT", quietly = TRUE)) {
  wide |>
    mutate(across(c(neurite_soma_ratio, neurite_fraction), \(x) round(x, 3))) |>
    DT::datatable(
      filter  = "top",
      rownames = FALSE,
      options = list(pageLength = 10, scrollX = TRUE)
    )
} else {
  knitr::kable(wide, digits = 3)
}
```

### Per-gene summary

```{r}
#| label: summarise

wide |>
  summarise(
    n              = n(),
    mean_ratio     = mean(neurite_soma_ratio, na.rm = TRUE),
    sd_ratio       = sd(neurite_soma_ratio,   na.rm = TRUE),
    median_ratio   = median(neurite_soma_ratio, na.rm = TRUE),
    .by = gene
  ) |>
  arrange(desc(mean_ratio)) |>
  knitr::kable(digits = 3)
```

### Ratio by gene

```{r}
#| label: fig-ratio
#| fig-cap: "Neurite:soma spot-count ratio. Each point is one replicate."
#| fig-width: 7
#| fig-height: 4.5

wide |>
  filter(!is.na(neurite_soma_ratio)) |>
  ggplot(aes(x = fct_reorder(gene, neurite_soma_ratio), y = neurite_soma_ratio)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, fill = "grey92") +
  geom_jitter(aes(colour = factor(date)), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_colour_brewer(palette = "Dark2", name = "Date") +
  labs(x = NULL, y = "Neurite / soma spot count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

## Export

```{r}
#| label: export
#| eval: false

readr::write_csv(wide, "pdf_mask_summary_wide.csv")
```

------------------------------------------------------------------------

::: callout-note
## Rendering

``` bash
# dummy data (default)
quarto render smFISH_pdf-mask_summary-report.qmd

# real data
quarto render smFISH_pdf-mask_summary-report.qmd \
  -P use_dummy_data:false \
  -P summary_csv:/path/to/pdf_mask_summary.csv
```
:::
