---
title: "HIV transcription foci and nucleus FISH intensities: plots"
format: html
editor_options:
  chunk_output_type: inline
---

Plots of the tables written by `measure-foci-nucleus-fish-intensities.ipynb` (section 2). Column names are converted to snake_case on loading (`janitor::clean_names()`), e.g. `total-foci-fish-normalised-integrated-intensity-relative-to-reference` becomes `total_foci_fish_normalised_integrated_intensity_relative_to_reference`.

## Environment

```{r}
#| label: setup
library(tidyverse)
library(ggbeeswarm)
library(rstatix)
library(ggpubr)
```

## Load tables

`replicate` and `field` are read as text so their leading zeros are kept (`080623`, `001`).

```{r}
#| label: load
out_dir <- "/Volumes/Jeff-exFAT/Marko_HIV/Marko_HIV_foci-quantification"

read_quant <- function(file, text_cols = c("replicate", "field")) {
  path <- file.path(out_dir, paste0("foci_intensity_quantification", file))
  header <- names(read_csv(path, n_max = 0, show_col_types = FALSE))
  text <- intersect(text_cols, header)
  read_csv(path,
           col_types = cols(.default = col_guess(),
                            !!!set_names(rep(list(col_character()), length(text)), text)),
           guess_max = Inf) |>
    janitor::clean_names()
}

set_levels <- function(df) {
  df |>
    mutate(across(any_of("rbp"), \(x) factor(x, levels = c("GFP", "CPEB4", "MARF1"))),
           across(any_of("induction"), \(x) factor(x, levels = c("NoDox", "Dox"))),
           across(any_of("hiv_infection"), \(x) factor(x, levels = c("Mock", "INF-24hpi", "INF-48hpi"))))
}

foci <- read_quant(".csv") |> set_levels()                               # one row per detected focus + one per nucleus without foci
nuclei <- read_quant("_summarised-per-nucleus.csv") |> set_levels()      # one row per nucleus (counted foci only)
mock_baseline <- read_quant("_mock-baseline.csv")                         # Mock background per pixel, per baseline group
mock_foci_filter <- read_quant("_mock-foci-filter.csv")                   # Mock foci cutoff and false-positive rate
reference_factors <- if (file.exists(file.path(out_dir, "foci_intensity_quantification_reference-factors.csv"))) {
  read_quant("_reference-factors.csv") |> set_levels()                   # per-replicate reference means (if REFERENCE_CONDITION was set)
}
```

## Design check

Nuclei per condition and replicate.

```{r}
#| label: design
nuclei |>
  count(rbp, induction, hiv_infection, replicate) |>
  pivot_wider(names_from = replicate, values_from = n)
```

## Plots

```{r}
#| label: plots

## foci count

nuclei %>%
  filter(foci_count > 0) %>%
  ggplot(aes(x = induction, y = foci_count, 
    colour = replicate, fill = replicate)) +
  geom_quasirandom(stroke = 0, alpha = 0.5) +
  facet_wrap(~rbp+hiv_infection) +
  theme_bw() 

View(foci)

## foci total FISH intensity

nuclei %>%
  filter(foci_count > 0) %>%
  ggplot(aes(x = induction, y = total_foci_fish_normalised_integrated_intensity_relative_to_reference)) +
  geom_quasirandom(
    aes(colour = replicate, fill = replicate),
    stroke = 0, alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~rbp+hiv_infection) +
  theme_bw() 


## foci mean FISH intensity per focus

nuclei %>%
  filter(foci_count > 0) %>%
  ggplot(aes(x = induction, y = mean_fish_normalised_integrated_intensity_per_foci_relative_to_reference)) +
  geom_quasirandom(
    aes(colour = replicate, fill = replicate),
    stroke = 0, alpha = 0.5) +
  geom_boxplot(width = 0.3) +
  scale_y_log10() +
  facet_wrap(~rbp+hiv_infection) +
  theme_bw() 


## nucleus FISH intensity 

nuclei %>%
  filter(foci_count > 0) %>%
  ggplot(aes(x = induction, y = nucleus_fish_normalised_integrated_intensity_relative_to_reference, 
    colour = replicate, fill = replicate)) +
  geom_quasirandom(stroke = 0, alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~rbp+hiv_infection) +
  theme_bw() 




```

## Statistics

Nuclei from the same replicate are not independent, so the replicate is the experimental unit (a "SuperPlot", Lord et al. 2020 JCB). Each replicate is reduced to one value per condition: the geometric mean (mean of log10) for intensities, the arithmetic mean for counts. Dox and NoDox are then compared with a paired t-test across replicates (pairs = same replicate), on log10 values for intensities. The test only runs for an `rbp` × `hiv_infection` panel with at least `min_pairs` replicates that have both Dox and NoDox. Holm-adjusted p-values (across panels) are in `p.adj`.

A non-parametric test on replicate means is not used: with 3–4 pairs, a Wilcoxon signed-rank test cannot give p < 0.125 whatever the data.

```{r}
#| label: stats-helpers
replicate_means <- function(df, y, log = TRUE) {
  df |>
    filter(foci_count > 0, if (log) .data[[y]] > 0 else TRUE) |>
    group_by(rbp, hiv_infection, induction, replicate) |>
    summarise(n_nuclei = n(),
              value = if (log) 10^mean(log10(.data[[y]])) else mean(.data[[y]]),
              .groups = "drop")
}

dox_test <- function(reps, log = TRUE, min_pairs = 3) {
  reps |>
    mutate(value = if (log) log10(value) else value) |>
    group_by(rbp, hiv_infection, replicate) |>
    filter(n_distinct(induction) == 2) |>                 # replicates with both Dox and NoDox
    group_by(rbp, hiv_infection) |>
    filter(n_distinct(replicate) >= min_pairs) |>
    arrange(induction, replicate, .by_group = TRUE) |>    # paired t-test pairs by row order
    t_test(value ~ induction, paired = TRUE) |>
    adjust_pvalue(method = "holm") |>
    add_significance("p.adj")
}

# Nuclei as small points, replicate means as large outlined points joined by replicate,
# Dox vs NoDox p-value (paired t-test on replicate means) where both are present.
superplot <- function(y, log = TRUE, ylab = y, label = "p = {p}") {
  cells <- nuclei |> filter(foci_count > 0, if (log) .data[[y]] > 0 else TRUE)
  reps <- replicate_means(nuclei, y, log)
  stats <- dox_test(reps, log) |>
    left_join(cells |>
                group_by(rbp, hiv_infection) |>
                # stat_pvalue_manual() takes y.position in log10 units on a log10 axis
                summarise(y.position = if (log) log10(max(.data[[y]])) + 0.15 else max(.data[[y]]) * 1.08,
                          .groups = "drop"),
              by = c("rbp", "hiv_infection"))
  print(stats)

  p <- ggplot(cells, aes(x = induction, y = .data[[y]], colour = replicate)) +
    geom_quasirandom(stroke = 0, alpha = 0.3, size = 1) +
    geom_line(data = filter(reps, n() == 2, .by = c(rbp, hiv_infection, replicate)),
              aes(y = value, group = replicate)) +
    geom_point(data = reps, aes(y = value, fill = replicate),
               shape = 21, size = 4, colour = "black") +
    stat_pvalue_manual(stats, label = label, tip.length = 0.01) +
    facet_wrap(~ rbp + hiv_infection) +
    labs(y = ylab) +
    theme_bw()
  expand <- expansion(mult = c(0.05, 0.12))
  if (log) p + scale_y_log10(expand = expand) else p + scale_y_continuous(expand = expand)
}
```

## SuperPlots

```{r}
#| label: superplots
superplot("foci_count", log = FALSE) + 
labs(title = "transcription foci count per nucleus", subtitle = "0-count removed, t-test comparison of replicate means") + 
theme_bw(base_size = 15)

superplot ("total_foci_fish_normalised_integrated_intensity_relative_to_reference") +
labs(title = "total transcription foci intensity per nucleus", subtitle = "0-count removed, t-test comparison of replicate means") + 
theme_bw(base_size = 15)

superplot("mean_fish_normalised_integrated_intensity_per_foci_relative_to_reference")

# nuclei with intensity <= 0 are dropped from the log axis and the geometric mean
superplot("nucleus_fish_normalised_integrated_intensity_relative_to_reference") +
labs(title = "nuclear HIV FISH intensity", subtitle = "nucleus only with one-or-more txn focus, t-test comparison of replicate means") + 
theme_bw(base_size = 15)
```
